{\bf {\Large Filtering time series of dissolved oxygen for improved estimates of estuary metabolism}}

### 2014-09-30 assess.Rnw
Finished mcm revisions round two, including more careful description of problem/potential solutions in intro

Results from the simulated time series were used to inform the validation of weighted regression with real data, specifically regarding half-window widths described below.  Continuous monitoring data from the \acl{NERRS} was used to validate the weighted regression model by evaluating estimates of ecosytem metabolism obtained from observed and filtered \ac{DO} time series. \ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \aclu{SWMP} (\acs{SWMP}, \citetalias{CDMO14}).  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.  We selected sites from the \ac{SWMP} database that had desirable characteristics for validating weighted regression.  Specifically, four macrotidal sites were chosen based on apparent relationships between \ac{DO} and tidal changes (\cref{fig:case_map,tab:case_att}): Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).

@@

An example from Sapelo Island illustrates the effects of weighted regression on \ac{DO} and metabolism estimates (\cref{fig:phase_out,fig:phase_in,fig:case_ex}).   A two-week period in February showed when the tidal cycles were both in and out of phase with the diel cycling, where phasing describes synchronicity between maximum tide heights and day/night periods.  That is, maximum tide heights were generally out of phase with the diel cycle during the first week when low tides were observed during the middle of the night and the middle of the day (\cref{fig:phase_out}), whereas tide heights were in phase during the second week when the maximum tide height occured during the day and night (\cref{fig:phase_in}).  The effects of tidal height change on the observed \ac{DO} time series are visually apparent in the plots. The first week illustrates a strong negative bias (less respiration, less production) in the observed \ac{DO} signal from low tides at mid-day and mid-night, whereas the second example illustrates a strong positive bias (more respiration, more production) in the observed \ac{DO} from high tides. These biases are apparent in the metabolism estimates using the observed data (\cref{fig:case_ex}).  Anomalous estimates occur when low tides are in phase with the solar cycle (week one), whereas metabolism estimates are likely over-estimated when high tides are in phase with the solar cycle (week two).  The filtered time series shows noticeable changes given the direction of bias from the phasing between tidal height and diel period.  \ac{DO} values were higher after filtering when low tides occurred during night and day periods, whereas \ac{DO} values were lower after filtering when high tides occurred during day and night periods (\cref{fig:phase_out,fig:phase_in}).  Changes in metabolism estimates after filtering were also apparent, such that the anomalous values were removed during the first week and the positive bias in the second week is decreased (\cref{fig:case_ex}).

@@

A point of concern is the period of observation within which observed \ac{DO} is affected by tidal height changes and the extent to which this affects the interpretation of ecosystem metabolism.  The effects of tidal variation on daily estimates may not be relevant if seasonal or annual aggregations remove this potential bias.  The example from Sapelo Island in the previous section highlights this point given that mean production and respiration estimates before and after filtering were generally unchanged for the two-week period. \cref{tab:case_res} also indicated that mean annual estimates of production and respiration were unchanged for Rookery Bay and Sapelo Island.  However, annual averages of production and respiration estimates were significantly different for Elkhorn Slough and Padilla Bay. Given these results, tidal variation may or may not have effects on metabolism estimates on time scales longer than 24 hours, depending on the location.  Therefore, an evaluation of weighted regression to filter the effects of tidal variation on ecosystem metabolism for different periods of observation is critical for its application.  Specifically, when should filtering be applied if aggregation of observed data on longer time periods removes potential bias?  A comparison of observed and filtered estimates that are aggregated over different periods of observation (e.g., annual, seasonal, monthly) could help address this question.

>>>>

Results from the simulated time series were used to inform the validation of weighted regression with real data, specifically with respect to choosing half-window widths described below.  Continuous monitoring data from the \acl{NERRS} was used to validate the weighted regression model by evaluating estimates of ecosytem metabolism obtained from observed and filtered \ac{DO} time series. \ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \aclu{SWMP} (\acs{SWMP}, \citetalias{CDMO14}).  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.  We selected sites from the \ac{SWMP} database that had desirable characteristics for validating weighted regression.  Specifically, four macrotidal sites were chosen based on apparent relationships between \ac{DO} and tidal changes (\cref{fig:case_map,tab:case_att}): Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).

@@

An example from Sapelo Island illustrates the effects of weighted regression on \ac{DO} and metabolism estimates (\cref{fig:phase_out,fig:phase_in,fig:case_ex}).   A two-week period in February showed when the tidal cycles were both in and out of phase with the diel cycling, where phasing describes synchronicity between maximum tide heights and day/night periods \citep{Nidzieko14}.  That is, maximum tide heights were generally out of phase with the diel cycle during the first week when low tides were observed during the middle of the night and the middle of the day (\cref{fig:phase_out}), whereas tide heights were in phase during the second week when the maximum tide height occured during the day and night (\cref{fig:phase_in}).  The effects of tidal height change on the observed \ac{DO} time series are visually apparent in the plots. The first week illustrates a strong negative bias (less respiration, less production) in the observed \ac{DO} signal from low tides at mid-day and mid-night, whereas the second example illustrates a strong positive bias (more respiration, more production) in the observed \ac{DO} from high tides. These biases are apparent in the metabolism estimates using the observed data (\cref{fig:case_ex}).  Anomalous estimates occur when low tides are in phase with the solar cycle (week one), whereas metabolism estimates are likely over-estimated when high tides are in phase with the solar cycle (week two).  The filtered time series shows noticeable changes given the direction of bias from the phasing between tidal height and diel period.  \ac{DO} values were higher after filtering when low tides occurred during night and day periods, whereas \ac{DO} values were lower after filtering when high tides occurred during day and night periods (\cref{fig:phase_out,fig:phase_in}).  Changes in metabolism estimates after filtering were also apparent, such that the anomalous values were removed during the first week and the positive bias in the second week is decreased (\cref{fig:case_ex}).

@@

A point of concern is the period of observation within which observed \ac{DO} is affected by tidal height changes and the extent to which this affects the interpretation of ecosystem metabolism.  The effects of tidal variation on daily estimates may not be relevant if seasonal or annual aggregations remove this potential bias.  The example from Sapelo Island in the previous section highlights this point given that mean production and respiration estimates before and after filtering were generally unchanged for the two-week period. \cref{tab:case_res} also indicated that mean annual estimates of production and respiration were unchanged for Rookery Bay and Sapelo Island.  However, annual averages of production and respiration estimates were significantly different for Elkhorn Slough and Padilla Bay. Given these results, tidal variation may or may not have effects on metabolism estimates on aggregated time scales longer than 24 hours, depending on the location.  Therefore, an evaluation of weighted regression to filter the effects of tidal variation on ecosystem metabolism for different periods of observation is critical for its application.  Specifically, when should filtering be applied if aggregation of observed data on longer time periods removes potential bias?  A comparison of observed and filtered estimates that are aggregated over different periods of observation (e.g., annual, seasonal, monthly) could help address this question.

### 2014-09-30 intro.Rnw
Finished mcm revisions round two, including more careful description of problem/potential solutions in intro

Open-water techniques have been used for decades to infer metabolic rates using \textit{in situ} measurements from continuous monitoring data \citep{Odum56}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  Originally conceived for streams \citep{Odum56}, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Russell07,Caffrey13}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  Such assumptions are often only implicity verified in practice, leading to potential biases.  The fundamental assumption is that the time series of \ac{DO} describes the same water mass over time \citep{Needoba12}.  Estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Russell07}.  Numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic given the potential effects of physical mixing \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}.

An extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  An important observation from this study regarding the utility of the open-water method was that a significant portion of the estimates were negative (3 - 69\% depending on site). Significant variation from advection of water masses with different metabolic histories was suggested as a likely factor influencing the \ac{DO} time series. Inaccurate air-sea exchange estimates may have also contributed to these seemingly anomalous values.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides. These studies provide compelling examples of the importance of potential noise in metabolism estimates related to physical advection.  The increasing availability of large, multi-annual datasets further warrants a need for quantiative methods that improve the accuracy of estimates of biological process rates.  \citet{Batt12} acknowledged this need by applying a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that account for cyclicity in time series associated with tidal variation in addition to process and observation uncertainty.

This article describes the application of a method for filtering an observed \ac{DO} time series for estimated tidal effects to more accurately quantify estimates of ecosystem metabolism for estuaries.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  Specifically, one year of \ac{DO} time series for each case study was filtered to adjust estimates of ecosystem metabolism to apparent tidal effects.  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous data along a tidal axis), we assume that tidal height is an appropriate and approximate measure that characterizes effects of physical advection.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.  Overall, the analysis is meant to better characterize the relative roles of biological and physical processes in estuarine systems.

>>>>

Open-water techniques have been used for decades to infer metabolic rates using \textit{in situ} measurements from continuous monitoring data \citep{Odum56}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  The fundamental assumption is that the time series of \ac{DO} describes the same water mass over time \citep{Needoba12}.  Estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Russell07,Caffrey13}.  Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Application of the method to estuaries is a particular concern as physical mixing caused by tidal currents may confound the biological variation in \ac{DO} time series \citep{Kemp80,Caffrey03,Nidzieko14}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in \ac{DO} signals from phyical advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting significant variation from advection of water masses with different metabolic histories was a likely factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of metabolism \citep{Murrell13}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  A control-volume approach was used by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} acknowledged this need by exploring the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the application of a method for filtering an observed \ac{DO} time series for estimated tidal effects to more accurately quantify estimates of ecosystem metabolism for estuaries.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  Specifically, one year of \ac{DO} time series for each case study was filtered to adjust estimates of ecosystem metabolism to apparent tidal effects.  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), we assume tidal height is an appropriate measure that approximates lateral variation.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.  Overall, the analysis is meant to better characterize the relative roles of biological and physical processes in estuarine systems.

### 2014-09-30 mat_proc.Rnw
Finished mcm revisions round two, including more careful description of problem/potential solutions in intro

The approach to filter physical advection from the observed \ac{DO} time series differs slightly from methods in \citet{Hirsch10}.  The previous approach used a two-dimensional grid predicted for stream pollutant concentrations across the time series and the range of discharge values observed in the study system \citep{Hirsch10}.  Normalized or discharge-independent values for pollutant concentration were obtained by averaging grid predictions across the discharge values that were likely to occur on a given day.  Rather than creating a two-dimensional grid of \ac{DO} related to time and tidal height change, the normalized time series herein were the model predictions conditional on time and constant tidal height set to the mean:
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}
such that the normalized time series represents \ac{DO} variation related to biological processes.  The term `filter' is used in reference to the removal of a specific variance component from the time series, while maintaining the structure of the biological component.  Although the approach shares similarities with common filtering techniques, a distinction is noted such that weighted regression has a specific purpose rather than more the generic objectives of common filters (e.g., moving window averages or local smoothers, \citealt{Shumway11}).

>>>>

The approach to filter physical advection from the observed \ac{DO} time series differs slightly from methods in \citet{Hirsch10}.  The previous approach used a two-dimensional grid predicted for stream pollutant concentrations across the time series and the range of discharge values observed in the study system \citep{Hirsch10}.  Normalized or discharge-independent values for pollutant concentration were obtained by averaging grid predictions across the discharge values that were likely to occur on a given day.  Rather than creating a two-dimensional grid of \ac{DO} related to time and tidal height change, the normalized time series herein were the model predictions conditional on time and constant tidal height set to the mean:
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}
such that the normalized time series represents \ac{DO} variation related to biological processes.  The term `filter' is used in reference to the removal of a specific variance component from the time series, while maintaining the structure of the biological component.  Although the approach shares similarities with common filtering techniques, a distinction is noted such that weighted regression has a specific purpose rather than the more generic objectives of common filters (e.g., moving window averages or local smoothers, \citealt{Shumway11}).
